---
title: "MIS 545 Project"
subtitle: "The Olympics"
author: "R. Straight"
date: "`r Sys.Date()`"
output:
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_html: default
bibliography: ["references.bib", "r-references.bib"]
biblio-style: "apalike"
link-citations: yes
---

```{r setup, include=FALSE}
library(citr)
library(tufte)
library(here)
library(tidyverse)
library(magrittr)
library(pander)
library(knitr)
library(janitor)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
here::here()
```

```{r source-all-data, include=FALSE}
# Load the data
athlete_events <- read_csv(here("data", "athlete_events.csv"))
noc_regions <- read_csv(here("data", "noc_regions.csv"))

# Source munging scripts
sys.source(here("munge", "01-A-loading_and_tidying.R"), envir = knitr::knit_global())
sys.source(here("munge", "02-A-Summarising.R"), envir = knitr::knit_global())
sys.source(here("munge", "03-A-Top_ten.R"), envir = knitr::knit_global())
```

# Introduction

The Olympics has become a financial juggernaut from the viewpoint of both the hosting country and the value of being a medalist. In 2016, the Rio Olymipcs cost $13.1 billion US dollars (7.23 billion reals) to host.^[[Rio Olympics cost \$13.1 billion](https://olympics.nbcsports.com/2017/06/14/rio-olympics-cost-13-billion/)]. This included "a subway line, a doping laboratory, a renovated port and cleanup of pouted Guanabara Bay." 

The value of _gold_ is more than the value of the medal, of course. Countries reward their medalists depending on which medal they bring home and these rewards vary drastically from country to country. Singapore, for example, rewards gold medalists with $1 million USD while Canada pays a comparatively paltry \$15,000 USD^[[Here’s how much Olympic athletes earn in 12 different countries](https://www.cnbc.com/2018/02/23/heres-how-much-olympic-athletes-earn-in-12-different-countries.html)]. Advertising sales during the Rio Olympics in 2016 totaled \$1.2 billion USD.^[[NBC says it has topped \$1 billion in national ad sales for 2020 Summer Olympics](https://www.cnbc.com/2019/12/10/nbc-says-it-topped-1b-in-national-ad-sales-for-2020-summer-olympics.html)]

`r newthought('Needless to say,')` it behooves interested parties to be able to predict just who, when, and where medalists will crop up, whether this is in an effort to determine if the 12-year-old male gymnast in Sweden is likely to be a 16-year-old gold medalist in four years or if this or that country is worth scouting in for talent given their past medal winnings.
 


# The Chosen Data

The purpose of this project is to apply a variety of classification and predictive methodologies to a chosen data set for the purposes of demonstrating knowledge and skills developed throughout the semester. The dataset chosen for this project is [120 years of Olympic history: athletes and results](https://www.kaggle.com/heesoo37/120-years-of-olympic-history-athletes-and-results)^[Described by the creator as "basic bio data on athletes and medal results from Athens 1896 to Rio 2016."] This particular dataset was chosen for a variety of reasons:

+ It is relatively large, coming in 271,116 rows when loaded raw.
+ There is a variety of variable types to work with, providing a range of options when it comes to different classification and preditive tests.
+ It affords a certain level of approachability and familiarity by virtue of its content; after all, we all know Olympic medalists.

`r newthought('The purpose of this')` study, then, is to examine the particulars of the Olympic historical record and attempt to identify trends and make predictions thereby. Three possibilities for this data in this context come to mind:

1. What trends are apparent in nations' medal totals?
1. What demographics contribute to medaling?
1. Can we predict a medal based on a collection of an athlete's demographics?

## A Description of the Data

The data originates in a [Kaggle.com dataset](https://www.kaggle.com/heesoo37/120-years-of-olympic-history-athletes-and-results?select=athlete_events.csv) provided by [Randi Griffin](https://www.kaggle.com/heesoo37). According to Griffin,

>This is a historical dataset on the modern Olympic Games, including all the Games from Athens 1896 to Rio 2016 [scraped from] www.sports-reference.com in May 2018.... Note that the Winter and Summer Games were held in the same year up until 1992. After that, they staggered them such that Winter Games occur on a four year cycle starting with 1994, then Summer in 1996, then Winter in 1998, and so on. A common mistake people make when analyzing this data is to assume that the Summer and Winter Games have always been staggered.
>
> `r tufte::quote_footer('--- Randi Griffin')`

The dataset is delivered in two files: `athlete_events.csv` and `noc_regions.csv`. The descriptions are provided in the data source.

::: {.fullwidth}
| File               	| Variable 	| Data type 	| Data format 	| Description                                               	|
|--------------------	|----------	|-----------	|-------------	|-----------------------------------------------------------	|
| athlete_events.csv 	|          	|           	|             	|                                                           	|
|                    	| ID       	| ind       	| int         	| Unique number for each athlete                            	|
|                    	| Name     	| ind       	| chr         	| Athlete's name                                            	|
|                    	| Sex      	| dep       	| chr         	| M or F                                                    	|
|                    	| Age      	| ind       	| int         	| Integer                                                   	|
|                    	| Height   	| ind       	| int         	| In centimeters                                            	|
|                    	| Weight   	| ind       	| num         	| In kilograms                                              	|
|                    	| Team     	| ind       	| chr         	| Team name                                                 	|
|                    	| NOC      	| ind       	| chr         	| National Olympic Committee 3 letter code                  	|
|                    	| Games    	| ind       	| chr         	| Year and season                                           	|
|                    	| Year     	| ind       	| int         	| Integer                                                   	|
|                    	| Season   	| ind       	| chr         	| Summer or Winter                                          	|
|                    	| City     	| ind       	| chr         	| City                                                      	|
|                    	| Sport    	| ind       	| chr         	| Sport                                                     	|
|                    	| Event    	| ind       	| chr         	| Event                                                     	|
|                    	| Medal    	| dep       	| chr         	| Gold, Silver, Bronze, or `NA`                             	|
| noc_regions.csv    	|          	|           	|             	|                                                           	|
|                    	| NOC      	| ind       	| chr         	| National Olympic Committee 3 letter code                  	|
|                    	| region   	| ind       	| chr         	| Country name (matches with regions in `map_data("world")` 	|
|                    	| notes    	| ind       	| chr         	| Notes                                                     	|
:::


# Data Pre-Processing

Fortunately, Griffin's scraping techniques prove tidy and in need of very little cleaning, all things considered. The entirety of the loading and tidying is as follows:

```{r
# tidy up the titles
athlete_events <- athlete_events %>% clean_names()
noc_regions <- noc_regions %>% clean_names()

# Join up athlete_events and noc_regions to get a nice country name
olympics <- as_tibble(athlete_events %>% left_join(noc_regions, by = "noc"))

# Switch to factors
olympics <- olympics %>%
  mutate(across(c("sex", "team", "noc", "games", "year", "sport", "medal", "city",
    "region", "season"), factor))

# Replace NAs in "medal" with "None"
olympics$medal <- olympics$medal %>% 
  as.character() %>% 
  replace_na("none") %>% 
  as_factor()

# There are way too many sports and a few only happened a couple times.
# Pare those down to the top 50, naming the rest "Other."
olympics$sport <- olympics$sport %>%
  fct_lump_n(n = 51)
```

With this we can now explore the data a bit easier. Note that each row in this dataset is for an _athlete_. Follows is a glimpse at the structure of the numerical data to verify data formats are as expected.


```{r opening-summary, echo=FALSE, message=FALSE}
library(kableExtra)
summary_table1 <- olympics %>% select(age, height, weight, year, medal) %>% summary()

summary_table1 %>% 
  kable(format = "latex", booktabs = T, caption = "Summary table of relevant data") %>% 
  kable_styling(latex_options = c("striped", "scaled_down"))
```

# Descriptive Analysis

```{r medals-by-year, echo=FALSE}
# Write upt eh code to get the total number of medals by country per games
```



# References

```{r bib, include=FALSE}
# create a bib file for the R packages used in this document
knitr::write_bib(c('base', 'rmarkdown'), file = 'r-references.bib')
```
